From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-27 Pi/paper/equivalence.tex
Use uniform style for references

We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
by relating them with a truncation. These can be found in~\cref{app:leveltwo} and in the accompanying Agda code.

@@

First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
syntax for 1-combinators is given in~\Cref{fig:piplus}.

@@

Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
$\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\Cref{fig:pihat}.

@@

Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and back from $\UFin$ to $\PiHatLang$. Types in $\PiHatLang$
are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The 1-combinators in
$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops in each connected component, equivalent to
$\Aut[\Fin[n]]$. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to
words in $\Sn$ and then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
2-paths between loops in $\UFin$.

@@

The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
combinator, compared to the direct interpretation given using the big-step interpreter in~\Cref{subsec:denotational}. We
verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.

@@

Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing
it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:

>>>>

We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
by relating them with a truncation. These can be found in~\cref{app:leveltwo} and in the accompanying Agda code.

@@

First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
syntax for 1-combinators is given in~\cref{fig:piplus}.

@@

Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
$\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\cref{fig:pihat}.

@@

Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and back from $\UFin$ to $\PiHatLang$. Types in $\PiHatLang$
are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The 1-combinators in
$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops in each connected component, equivalent to
$\Aut[\Fin[n]]$. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to
words in $\Sn$ and then to automorphisms using~\cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
2-paths between loops in $\UFin$.

@@

The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
combinator, compared to the direct interpretation given using the big-step interpreter in~\cref{subsec:denotational}. We
verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.

@@

Recall the specification of reversible disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing
it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:

### 2021-10-27 Pi/paper/finite.tex
Use uniform style for references

% Recap of the previous section
In~\Cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
is also known to be the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$
for every $n$. In this section, we will describe this group syntactically.

@@

\begin{proposition}
  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the empty list $\nil$ for the neutral
  element, multiplication given by list append $\append$, and inverse given by flipping $\inl$ and $\inr$, followed by
  reversing the list. Further, $F(A)$ with $\eta_A(a) \defeq \inl(a) \cons \nil$ satisfies the universal property of
  free groups, as stated in~\Cref{prop:free-groups}.
\end{proposition}

@@

The universal property of the above definition is similar to~\Cref{prop:free-groups} except the relation has to be
preserved by the function mapping out of the generating set.

@@

\paragraph{Coxeter Presentation} To present the group $\Sn$, the primitive operations we use will be adjacent swaps.
When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions -- transposition number
$k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
going to specify for this presentation -- we visualise them as braid diagrams in~\Cref{fig:coxeter-braid}.

@@

For this to work, we first need the system to have the termination property, meaning that there are no infinite
reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can get a normalisation function. This will be true if the
rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
rules, have to converge. For example, in our system, the pair in~\Cref{fig:critical-pairs-converging} converges.
Unfortunately, this is not true for all critical pairs -- an  example is in~\Cref{fig:critical-pairs-non-converging},
where left and right endpoints are normal with respect to the $\cox*$ relation.

@@

By~\Cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
  {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state two important properties enjoyed
by $\normf$.

@@

\noindent
Notice however, that for a group presentation, as defined in~\Cref{def:presentation}, the relation needs to be on the
set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
proposition lets us lift a group structure on $\List[A]$ to a group presentation.

@@

\noindent By composing~\Cref{prop:lehmer-aut-equiv} and~\Cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.

>>>>

% Recap of the previous section
In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
is also known to be the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$
for every $n$. In this section, we will describe this group syntactically.

@@

\begin{proposition}
  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the empty list $\nil$ for the neutral
  element, multiplication given by list append $\append$, and inverse given by flipping $\inl$ and $\inr$, followed by
  reversing the list. Further, $F(A)$ with $\eta_A(a) \defeq \inl(a) \cons \nil$ satisfies the universal property of
  free groups, as stated in~\cref{prop:free-groups}.
\end{proposition}

@@

The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
preserved by the function mapping out of the generating set.

@@

\paragraph{Coxeter Presentation} To present the group $\Sn$, the primitive operations we use will be adjacent swaps.
When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions -- transposition number
$k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
going to specify for this presentation -- we visualise them as braid diagrams in~\cref{fig:coxeter-braid}.

@@

For this to work, we first need the system to have the termination property, meaning that there are no infinite
reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can get a normalisation function. This will be true if the
rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
rules, have to converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges.
Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
where left and right endpoints are normal with respect to the $\cox*$ relation.

@@

By~\cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
  {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state two important properties enjoyed
by $\normf$.

@@

\noindent
Notice however, that for a group presentation, as defined in~\cref{def:presentation}, the relation needs to be on the
set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
proposition lets us lift a group structure on $\List[A]$ to a group presentation.

@@

\noindent By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.

### 2021-10-27 Pi/paper/introduction.tex
Use uniform style for references

      \item We start in~\Cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to

@@

      \item In~\Cref{sec:pi}, we introduce the two-level language

@@

      \item ~\Cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in

@@

      \item In~\Cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group

@@

      \item In~\Cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in

@@

      \item In~\Cref{sec:applications}, we show applications of our results to reversible circuits, using our

>>>>

      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to

@@

      \item In~\cref{sec:pi}, we introduce the two-level language

@@

      \item ~\cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in

@@

      \item In~\cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group

@@

      \item In~\cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in

@@

      \item In~\cref{sec:applications}, we show applications of our results to reversible circuits, using our
